# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

1) Load the data

```{r load,results='hide'}
#Load activity file
unzip(zipfile = "activity.zip",files = "activity.csv")
#Read 'date' variable as strings instead of converting as factors
activity_monitoring <- read.csv("activity.csv",stringsAsFactors=FALSE)
#Remove input file from /tmp once reading is done
file.remove("activity.csv")
```

2) Process/transform data(if necessary) into a format suitable for analysis

one simple transformation is converting "date" variable from character to date datatype

```{r preproc}
#Converting 'date' variable from character to date
activity_monitoring$date <- as.Date(activity_monitoring$date)
#Checking summary of input data to understand data
summary(activity_monitoring)
#Observation : 2304 missing values in "steps"
```

## What is mean total number of steps taken per day?

1) Histogram of the total number of steps taken each day

```{r totsteps_hist}
#Ignoring missing values.
total_steps_per_day <- aggregate(steps ~ date, data=activity_monitoring, sum,na.rm = T)
hist(total_steps_per_day$steps,main="Histogram of total number of steps taken each day", xlab="Steps", col="blue")
```


2) Calculating mean of the total steps taken per day. Median calculation follows next.

```{r totsteps_mean}
#Ignoring missing values. calculate mean of the total steps taken per day
mean(total_steps_per_day$steps,na.rm=T)
```

2) Calculating median of the total steps taken per day

```{r totsteps_median}
#Ignoring missing values. calculate median of the total steps taken per day
median(total_steps_per_day$steps,na.rm=T)
```


## What is the average daily activity pattern?

1) A time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseriesplot1}
#Aggregate to find average number of steps taken by interval
aggr_by_interval <- aggregate(steps ~ interval,data=activity_monitoring, mean,na.rm = T)
#Time Series Plot
with(aggr_by_interval, plot(steps~interval, type="l", xlab="Interval", ylab="Avg number of steps taken", main="Time series plot", col="red"))
```

2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r intervalmaxsteps}
max_steps <- aggr_by_interval[which.max(aggr_by_interval$steps),]
print("Following interval has maximum number of averaging steps by interval across all the days in dataset")
print(max_steps$interval)
```


## Imputing missing values

1) Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r totalNAs}
sapply(X = activity_monitoring,function (x) { sum(is.na(x))})
#There are about 2304 missing values in steps
```

2) Devising a strategy for filling in all of the missing values in the dataset. 

I've chosen mean for 5 minute interval and impute on the missing values. 

```{r impute}
#Using arrange & ddply functions from plyr
library(plyr)

#Imputing missing values in "steps" - New variable called "imputed_steps" is added without disturbing
#existing column. This is helpful to check quality of imputed values. Imputed_steps attribute will
#be used for all further analysis

temp_df <- ddply(activity_monitoring, "interval", transform, imputed_steps=ifelse(is.na(steps), round(mean(steps,na.rm=T)), steps))

#Sort data frame by interval and date to maintain in the original format order
imputed_activity_monitoring <- arrange(temp_df, date, interval)
#Dropping temp dataframe
rm(temp_df)
```

3) New data set has been created already with imputed missing values using above strategy [mean for 5 minute interval]. Just displaying header rows to check the data set

```{r imputeddataset}
head(imputed_activity_monitoring)
```

4) Making a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r imputed_totalstep_hist}
#wanted to use ddply instead of aggregate to get total number of steps taken each day
imputed_total_steps_per_day <- ddply(imputed_activity_monitoring,"date",summarise,tot_steps=sum(imputed_steps))
hist(imputed_total_steps_per_day$tot_steps,main="Histogram of total number of steps taken each day(after imputing)", xlab="Steps", col="orange")
```

Additionaly, I am creating Side by side comparison of histogram which should yield good interpreation of histograms

```{r compare_histogram}
par(mfcol = c(1,2))
hist(total_steps_per_day$steps,main="Histogram of total number of \nsteps taken each day\n(Before imputing)", xlab="Steps", col="blue")
hist(imputed_total_steps_per_day$tot_steps,main="Histogram of total number of \nsteps taken each day\n(after imputing)", xlab="Steps", col="orange")

par(mfcol = c(1,1))
```

Calculating mean of the total steps taken per day

```{r imputed_totsteps_mean}

mean(imputed_total_steps_per_day$tot_steps)
```

calculating medain of the total steps taken per day

```{r imputed_totsteps_median}

median(imputed_total_steps_per_day$tot_steps)
```

**Results:** *From the above observations, imputed values did not affect shape of histogram much, but there is a little difference in height of the histogram. When there is no change in the mean, Median changed little bit. I am presuming that impute values did not disturb pattern of data much. I believe that is good*

## Are there differences in activity patterns between weekdays and weekends?

1) Creating a new factor variable "weekday_ind" with two levels - "Weekday" & "Weekend"
```{r diff_weekday_weekend}
#Adding weekday indicator with existing dataset
imputed_activity_monitoring <- ddply(imputed_activity_monitoring,"date",transform,weekday_ind= ifelse(weekdays(date) %in% c("Saturday","Sunday"),"Weekend","Weekday"))

#Averaging steps by weekday_ind and interval
avg_weekly_activity <- ddply(imputed_activity_monitoring, c("interval","weekday_ind"),summarise, avg_steps = mean(imputed_steps))

```

2) Making a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

My choice here is ggplot2 graphical system

```{r panelplot}
library(ggplot2)

ggplot(data=avg_weekly_activity, aes(x=interval,y=avg_steps, colour=weekday_ind), ) + geom_line() + facet_wrap(~weekday_ind,nrow=2) + xlab("Interval") + ylab("Average Steps Taken") + ggtitle("Panel plot comparison of average steps taken \nby Week days and Week ends")
```

**Results:** *My interpretation by comparing weekday and weekend plots - People are active mostly in weekends. I do see several average steps peeks in weekend after 1000 minutes interval. However, people are active in weekdays than weekend on specific interval, that is, between 750 to 900 minutes approximately*


